The Iterative Signature Algorithm

### 2009-04-28 isa/inst/doc/ISA_tutorial.Rnw
isa vignette update, almost finished. Some captions missing and a proofreading is missing as well.

@@

@@

>>>>

In this section we show some examples that highlight some key features
of the ISA: its resilience to noise and its ability to find
overlapping modules.

@@

To test the behavior of ISA with noisy inputs we generate a number of
in-silico data sets with different noise levels
(Fig.~\ref{fig:noise}).
<<noisegen,cache=TRUE>>=
noise <- seq(0.1,1,by=0.1)
data <- lapply(noise, function(x) isa.in.silico(500,200,10,noise=x))
@

<<noiseplot1,echo=FALSE,results=hide>>=
png("noise.png", width=800, height=320)
@ 
<<noiseplot2>>=
layout( matrix(1:10, nr=2, byrow=TRUE) )
for (i in seq_along(data)) {
  par(mar=c(1,1,1,1))
  image(1-data[[i]][[1]], axes=FALSE, frame=TRUE)
}
@ 
<<noiseplot3,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{noise}
  \caption{}
  \label{fig:noise}
\end{figure}

Next, we run ISA with the default parameters on all the data
sets. This might take a while.
<<noise.isa,cache=TRUE>>=
modules <- lapply(data, function(x) isa(x[[1]]))
@

Let us check the sensitivity of ISA, for every real module, we pick
one from the ones ISA has found, the one that is the most correlated
to it.
<<sensitivity.noise>>=
best <- lapply(seq_along(modules), function(i) {
  cc <- pmin(cor(modules[[i]]$rows!=0, data[[i]][[2]]),
             cor(modules[[i]]$columns!=0, data[[i]][[3]]))
  apply(cc, 2, max)
})
best.mean <- sapply(best, mean)
@

Let's create boxplots for the different noise levels.

<<sensbox,include=FALSE,results=hide>>=
boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<sensbox>>
@ 
\caption{}
\label{fig:sensbox}
\end{figure}

To test the specificity, we create boxplots for the correlation
coefficients of all modules found by ISA and their closest real
module.
<<spec>>=
spec <- lapply(seq_along(modules), function(i) {
  cc <- pmin(cor(modules[[i]]$rows != 0, data[[i]][[2]]),
             cor(modules[[i]]$columns != 0, data[[i]][[3]]))
  apply(cc, 1, max)
})
@ 
<<specbox,include=FALSE,results=hide>>=
boxplot(spec, names=noise, xlab="Noise level", ylab="Specificity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<specbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<specbox>>
@
\caption{}
\label{fig:specbox}
\end{figure}

@@

ISA biclusters might overlap in their rows, columns, or both. To
illustrate this, we create an artificial data set with two overlapping
blocks.

<<set.seed.overlap>>=
set.seed(1)
@ 
<<overlap>>=
two.over <- isa.in.silico(100,100,2,40,40,overlap.row=10)
@

We run ISA on this input matrix, with two threshold parameters, the
first thresholds are very mild, they are both zero.
In other words, in each iteration step all rows/columns are kept in each
that have a higher score than the average. The second set of
thresholds are stricter, the scores have to be at least one standard
deviation away from the mean to keep them.

<<overlap.isa>>=
ov.normed <- isa.normalize(two.over[[1]])
ov.seeds <- generate.seeds(count=100, length=100)
ov.modules.1 <- isa.iterate(ov.normed, ov.seeds, 
                            thr.row=0, thr.col=0, direction="up")
ov.modules.1 <- isa.unique(ov.normed, ov.modules.1)
ov.modules.1 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.1)
ov.modules.2 <- isa.iterate(ov.normed, ov.seeds, 
                            thr.row=1, thr.col=1, direction="up")
ov.modules.2 <- isa.unique(ov.normed, ov.modules.2)
ov.modules.2 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.2)
ncol(ov.modules.1$rows)
ncol(ov.modules.2$rows)
@

ISA found \Sexpr{ncol(ov.modules.1$rows)} %$
modules with the mild thresholds and \Sexpr{ncol(ov.modules.2$rows)} %$
modules with the strict ones. Let us plot the original data and the
modules found by ISA (Fig.~\ref{fig:overlapplot}).

<<overlapplot,include=FALSE,results=hide,cache=TRUE>>=
layout( rbind(1:4, c(8,5,6,7)) )
par(mar=c(1,1,3,1))
image(2-two.over[[1]], axes=FALSE, frame=TRUE, main="Input matrix")
for (i in 1:3) {
  par(mar=c(1,1,3,1))
  image(1-outer(ov.modules.1$rows[,i]!=0, ov.modules.1$columns[,i]!=0),
        axes=FALSE, frame=TRUE, main=paste("Module", i))
}
for (i in 1:3) { 
  par(mar=c(1,1,3,1))
  image( 1-outer(ov.modules.2$rows[,i]!=0, ov.modules.2$columns[,i]!=0),
        axes=FALSE, frame=TRUE, main=paste("Module", i+3))
}
@

\begin{figure}
\centering
<<overlapplot2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=8,height=5>>=
<<overlapplot>>
@ 
\caption{}
\label{fig:overlapplot}
\end{figure}

As expected the ISA modules are in general bigger with the mild
thresholds. ISA correctly identifies the two overlapping modules and
also the union of them. With the stricter thresholds it find the
non-overlapping parts of the two modules, plus their overlap as a
separate module.

\section{Running ISA in parallel}

Since a typical ISA analysis consists of using a range of row/column
and these runs are independent of each other; it is trivial to
parallelize the task by performing the iterations for different
threshold parameters on different processors or computers. Here we
show and example on how to do this easily with the \Rpackage{snow} and
the \Rpackage{Rmpi} packages:
<<parallelloadpackage>>=
library(snow)
library(Rmpi)
@

We generate some simple in-silico data.
<<paralleldata>>=
pdata <- isa.in.silico()
@

Create the MPI cluster, with eight working nodes. You need to have a
working MPI installation for this. See more in the documentation of
the \Rpackage{Rmpi} and \Rpackage{snow} packages.
<<createcluster>>=
clu <- makeMPIcluster(8)
invisible(clusterEvalQ(clu, library(isa)))
clusterExport(clu, "pdata")
@

Create a big matrix in which each row is a combination of the
threshold parameters.
<<parthr>>=
thr <- seq(1,3,by=0.2)
thr.list <- expand.grid(thr, thr)
@

First we run the ISA on a single processor only, and measure the
running time.
<<seqrun,cache=TRUE>>=
system.time(modules <- isa(pdata[[1]], thr.row=thr, thr.col=thr))
@

Let us now do a parallel run, again, with measuring the running
time. If you are really running this on multiple CPUs, then it is much
faster.
<<parrun,cache=TRUE>>=
system.time(modules.par <- parApply(clu, thr.list, 1, function(x) {
  isa(pdata[[1]], thr.row=x[1], thr.col=x[2])
}))
@

Finally, stop the cluster.
<<closecluster>>=
stopCluster(clu)
@
